---
title: "Babynames Exercise"
author: "Fiona Fang"
output:
  html_document:
    toc: true
    self_contained: true
    keep_md: true
  ioslides_presentation:
    smaller: yes
    keep_md: true
  pdf_document:
    toc: true
    df_print: kable

font-family: Helvetica
subtitle: Data Visualization - Columbia University
autosize: yes
---

## Preparation

```{r setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library("knitr")
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, message=FALSE, warning = FALSE)
```

Install data package:

```{r, echo=TRUE}
library(babynames)
```

```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)    
library(tidyverse) 
```


## 1. Plot the most common names in 2017 over the entire period

### Select the top 10 boys and girls names in 2017 from the overall dataset

```{r}
common_names_2017 <- babynames %>% 
  filter(year == 2017) %>% 
  group_by(sex, name) %>%
  summarize(total = sum(n)) %>%
  arrange(desc(total)) %>%
  group_by(sex) %>%
  mutate(rank=row_number()) %>%
  filter(rank<=10)  %>% # top 10
  arrange(sex, rank)

common_names_2017
```

### Separate them into two groups: boys and girls

```{r}
common_names_2017f <- common_names_2017 %>% filter(sex=="F")
common_names_2017m <- common_names_2017 %>% filter(sex=="M")
```

### Top 10 common names in 2017 for girls over the entire period

```{r}
babynames %>%
  filter(sex=="F") %>% # for girls
  filter(name %in% common_names_2017f$name) %>%
  ggplot(., aes(year, n)) +
  geom_line(aes(color=name, group=name)) + 
  labs(title = "Trend of Top 10 Girl Names of 2017 Over Time", 
       x = "Year", y = "Number of Babies")
```

### Top 10 common names in 2017 for boys over the entire period

```{r}
babynames %>%
  filter(sex=="M") %>% # for boys
  filter(name %in% common_names_2017m$name) %>%
  ggplot(., aes(year, n)) +
  geom_line(aes(color=name, group=name)) + 
  labs(title = "Trend of Top 10 Boy Names of 2017 Over Time", 
       x = "Year", y = "Number of Babies")
```


## 2. Explore which names are most often used as unisex names. For which names has the popularity over time changed a lot?

### Top 10 common unisex names

Unisex: designed to be suitable for both sexes

```{r}
top10_unisex_names <- babynames %>%
  group_by(name) %>%
  filter(all(c("F", "M") %in% unique(sex))) %>% # unisex names
  summarize(total = sum(n)) %>%
  ungroup() %>%
  top_n(10, total) # Top 10

top10_unisex_names
```

```{r}
top_unisex_names_data <- babynames %>%
  filter(name %in% top10_unisex_names$name)
```

### Plot the data

```{r}
ggplot(top_unisex_names_data, aes(x = year, y = n, color = name)) +
  geom_line() +
  facet_wrap(~name) +
  labs(title = "Popularity of Top 10 Unisex Names Over Time",
       x = "Year",
       y = "Number of Babies Named")

```

The popularity of specific names like David, James, John, Mary, Michael, Robert changed a lot over time.

## 3. Identify one particular pattern in the data 

### Define a list of religious names

```{r}
religious_names <- c("Mary", "Joseph", "Jesus", "Moses", "Abraham", "Isaac", "Solomon", "Mohammed", "Krishna","Rama","Siddhartha")
```

```{r}
religious_names_data <- babynames %>%
  filter(name %in% religious_names)

yearly_totals <- religious_names_data %>%
  group_by(year) %>%
  summarize(total = sum(n)) %>%
  ungroup()
```

### Plot the trends

```{r}
ggplot(yearly_totals, aes(x = year, y = total)) +
  geom_line() +
  labs(title = "Trends in Popularity of Religious Names Over Time",
       x = "Year",
       y = "Total Number of Babies Named")
```

As the plot shows, religious names are more and more common from 1880 to 1920, and less common after 1960. This trend can reflect various cultural, social, and historical influences. 

